---
title: 'Jones_Clayton Data Analysis Assignment #1 (50 points total)'
author: "Jones, Clayton"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS,mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: SEX and CLASS are categorical data, whereas the rest of the data is all numerical. However, class is determined by numerical data (RINGS). By looking at the summary, it can be observed that the WHOLE,SHUCK and VOLUME have maximum values significantly higher than the mean and median values, possible indicating potential skewness to the right or outliers. Potential sknewness in the data could show correlation (but not necessarily causation) of how the abalones act as they age, how their physical profiles correlate to how they age, how sex correlates with size, and much more. ***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}

addmargins(table(mydata$SEX, mydata$CLASS, dnn = c("SEX","CLASS")))

barplot(table(mydata$SEX, mydata$CLASS),beside = TRUE, legend = c("Female","Infant","Male"),col = c("blue","green","red"),xlab = "Classiciation",ylab = "Frequency")

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: Looking at the graph above, it is clear that the lower classifications of abalones consists of more infants, meaning younger abalones (which makes sense). As we move into the higher classifications, the number of infants decrease and the number of males and females increase, with the number of males and females both peaking in the A3 classification. While there are more male abalones classified as A3 than females, the classifications beyond (A4 and A5) show a roughly even distribution of males and females.The distriution of abalones by class is slightly skewed to the left.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
indices<-sample(1:1036, size = 200)
work<-mydata[indices,]

plot(work[,2:6])
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$VOLUME,mydata$WHOLE, col = mydata$CLASS, ylab = "Whole weight (g)", xlab = "Volume", main = "Whole Weight vs Volume", pch = 16)
legend("topleft",legend = levels(mydata$CLASS), col = unique(mydata$CLASS), pch = 16)
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE,mydata$SHUCK, col = mydata$CLASS, ylab = "Shuck weight",xlab = "Whole weight", main = "Shuck weight vs whole weight",pch = 16)
legend("topleft",legend = levels(mydata$CLASS), col = unique(mydata$CLASS), pch = 16)
abline(a=0,b=max(mydata$SHUCK/mydata$WHOLE))

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: It appears as though there is less variability in the relationship of shuck weight vs whole weight than there is in whole weight vs volume. At lower whole weights, both shuck weight and volume show less variability but begin to show more variability as whole weight increases. There is more variability in volume as whole weight increases than shuck weight as it increases. It appears as though for the A3 classification, the ratio of shuck weight to whole weight is lower than the other classes. This is not the case for volume, as that is pretty variable relative to the other classifications. The graph indicates that in the younger classifications (A1 and A2), the shuck takes up more of the whole weight than it does as the abalone ages.  ***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata[mydata[,1] == "F", "RATIO"], col = "blue", xlab = "RATIO",main = "Female RATIO")
hist(mydata[mydata[,1] == "I", "RATIO"], col = "green", xlab = "RATIO",main = "Infant RATIO")
hist(mydata[mydata[,1] == "M", "RATIO"], col = "red", xlab = "RATIO",main = "Male RATIO")
boxplot(mydata[mydata[,1] == "F", "RATIO"], col = "blue", xlab = "RATIO",main = "Female RATIO")
boxplot(mydata[mydata[,1] == "I", "RATIO"], col = "green", xlab = "RATIO",main = "Infant RATIO")
boxplot(mydata[mydata[,1] == "M", "RATIO"], col = "red", xlab = "RATIO",main = "Male RATIO")
qqnorm(mydata[mydata[,1] == "F", "RATIO"], col = "blue", main = "Female RATIO")
qqline(mydata[mydata[,1] == "F", "RATIO"])
qqnorm(mydata[mydata[,1] == "I", "RATIO"], col = "green",main = "Infant RATIO")
qqline(mydata[mydata[,1] == "I", "RATIO"])
qqnorm(mydata[mydata[,1] == "M", "RATIO"], col = "red", main = "Male RATIO")
qqline(mydata[mydata[,1] == "M", "RATIO"])
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: By looking at the boxplots, it can be observed that the RATIO appears to be skewed to the right for all three sexes. The boxplots show that the RATIOs appear to be fairly normal with a few outliers at higher RATIOs. The QQ plots agree with the boxplots, that the distributions are fairly normal between the -2 and 2 theoretical quantiles. But beyond those quantiles, the data does not stick to the line, implying that the data is not normal. ***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
f_outliers<-boxplot.stats(mydata$RATIO[mydata$SEX == "F"])$out
i_outliers<-boxplot.stats(mydata$RATIO[mydata$SEX == "I"])$out
m_outliers<-boxplot.stats(mydata$RATIO[mydata$SEX == "M"])$out

outliers<-c(f_outliers,i_outliers,m_outliers)
kable(mydata[mydata$RATIO %in% outliers,])

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: For males and females, most outliers observed belong to class A3 whereas for infants, most outliers observed belong to class A1. While males and infants only had outliers that were larger than expected given the distribution, there was one female outlier that had a very small RATIo value. In general, outliers are more observable and common on the larger end of the RATIO spectrum.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow = c(2,2))
boxplot(mydata$VOLUME ~ mydata$CLASS, xlab = "Class", ylab = "Volume", col = unique(mydata$CLASS))
boxplot(mydata$WHOLE ~ mydata$CLASS, xlab = "Class", ylab = "Whole weight",col = unique(mydata$CLASS))
plot(mydata$RINGS, mydata$VOLUME, xlab = "Rings", ylab = "Volume", col = unique(mydata$CLASS), pch = 16)
legend("topright",inset = c(-.1,0),legend = levels(mydata$CLASS), col = unique(mydata$CLASS), pch = 16, xpd = TRUE, cex = .5)
plot(mydata$RINGS, mydata$WHOLE, xlab = "Rings", ylab = "Whole weight", col = unique(mydata$CLASS), pch = 16)
legend("topright",inset = c(-.1,0),legend = levels(mydata$CLASS), col = unique(mydata$CLASS), pch = 16, xpd = TRUE, cex = .5)
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Both variables of VOLUME and WHOLE are decent indicators of age. RINGS is based on age of the abalone, and because CLASS is based on RINGS, CLASS is also based on age. Because age increases with classification, the boxplots indicate that there is a positive correlation between age (CLASS) and both VOLUME and WHOLE. The same can be said for the scatter plots, as the graphs indicate a positive corrlation between age (RINGS) and both VOLUME and WHOLE.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
avg_vol<- aggregate(VOLUME ~ SEX + CLASS, data = mydata,mean)
avg_shuck<- aggregate(SHUCK ~ SEX + CLASS, data = mydata,mean)
avg_ratio<- aggregate(RATIO ~ SEX + CLASS, data = mydata,mean)


kable(matrix(round(avg_vol$VOLUME,2), nrow =3, byrow = FALSE,dimnames = list(unique(levels(mydata$SEX)),unique(levels(mydata$CLASS)))),caption = "Average VOLUME by SEX and CLASS")

kable(matrix(round(avg_shuck$SHUCK,2), nrow =3, byrow = FALSE,dimnames = list(unique(levels(mydata$SEX)),unique(levels(mydata$CLASS)))),caption = "Average SHUCK by SEX and CLASS")

kable(matrix(round(avg_ratio$RATIO,2), nrow =3, byrow = FALSE,dimnames = list(unique(levels(mydata$SEX)),unique(levels(mydata$CLASS)))),caption = "Average RATIO by SEX and CLASS")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
par(mfrow=c(1,3))
interaction.plot(x.factor = avg_ratio$CLASS,trace.factor = avg_ratio$SEX, response = avg_ratio$RATIO, col = unique(avg_ratio$SEX), xlab = "CLASS", ylab = "Average RATIO")
interaction.plot(x.factor = avg_vol$CLASS,trace.factor = avg_vol$SEX, response = avg_vol$VOLUME, col = unique(avg_vol$SEX), xlab = "CLASS", ylab = "Average VOLUME")
interaction.plot(x.factor = avg_shuck$CLASS,trace.factor = avg_shuck$SEX, response = avg_shuck$SHUCK, col = unique(avg_shuck$SEX), xlab = "CLASS", ylab = "Average SHUCK")

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: These plots raise a a number of different questions. First, it can be observed that by looking at the graph of avergae RATIO vs CLASS, there is an inverse relationship, meaning RATIO decreases as the abalones age. However, by looking at the next two graphs, both SHUCK and VOLUME increase as the abalones age, and it appears as though at the same rate. However, the slopes are not the same, so this would make sense. But one question would be is it typical that as abalones age, their VOLUME is more spread out to their areas besides their SHUCK? It can also be seen that mean SHUCK weight and VOLUME peak in class A4. One question would be why do these things decrease from A4 to A5 as they age? Are they like humans in the fact that as they really start to age, they begin to shrink? ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow = c(2, 2))

infantdata<-subset(mydata, RINGS < 16 & SEX == "I")
adultdata<-subset(mydata, RINGS < 16 & SEX != "I")
boxplot(infantdata$VOLUME ~ infantdata$RINGS, col = "green", ylim = c(0,1100), xlab = "RINGS", ylab = "VOLUME", main = "Infant VOLUME by RINGS")
boxplot(adultdata$VOLUME ~ adultdata$RINGS, col = "purple", ylim = c(0,1100), xlab = "RINGS", ylab = "VOLUME", main = "ADULT VOLUME by RINGS")
boxplot(infantdata$WHOLE ~ infantdata$RINGS, col = "green", ylim = c(0,400), xlab = "RINGS", ylab = "WHOLE", main = "Infant WHOLE by RINGS")
boxplot(adultdata$WHOLE ~ adultdata$RINGS, col = "purple", ylim = c(0,400), xlab = "RINGS", ylab = "WHOLE", main = "Adult WHOLE by RINGS")

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays suggest that VOLUME and WHOLE increase in infant abalaones until they hit a certain age (RINGS value ~ 10) and then stop growing. The same goes for adult abalones, but the RINGS value is approximately 11. The displays also show that adult abalones have much more variability in terms of VOLUME and WHOLE as they age than infant abalones.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: There are a number of different statistical reasons that explain the failure of this study. First, one thing explaining the failure of the original study could be how the abalones are determined to be infants. This can be a difficult process,  but some of the data shows infants with physical characteristics similar to the oldest males and females. This doesn't really make sense and can make the data funky. Also, because ring clarity can be an issue, it is difficult to determine the age of the abalones and it is a very time consuming process. RINGS and CLASS are both dependent on this field being accurate, so if it is not correct, it can cause problems in the data analysis and potentially lead to inaccurate conclusions. Also, some of the data categories here like VOLUME and RATIO are determined from other characteristics in the data set. Because this is the case, it can be hard to draw conclusions because those characteristics are dependent on other characteristics. There also seems to be some skewness within the data, meaning it is possible that the sample size is not large enough. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: There are a number of different questions I would ask if I were presented with an overall histogram and summary statistics but no other information. One would be how large is the sample size? How was the sample selected? I would also like to know if there are any outliers or any outstanding conditions that could cause skewness in the data.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: It may not always be the case, but data from observational studies could contain human error in values that have to be measured in order to be recorded. Observational studies can also contain biases, such as selection bias. Confounders can also be a problem with observational studies, a variable outside the scope of the study that could have an impact on the measurements. Causality cannot be determined because the studies are just observational, meaning nothing is being manipulated or tested.These observational studies can show relationships and correlations between variables that can lead to studies to try and prove causality, but they themselves cannot prove causality.***
